---
title: "Differential Gene Expression"
author: "Simon Haugaard"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Required R libraries
```{r library}
if (!require("pacman")) install.packages("pacman")
pacman::p_load("edgeR", "readr", "readxl", "biomaRt", "magrittr", "tibble", "stringr", 
               "ggplot2", "data.table", "patchwork", "openxlsx", "aamisc", "limma", 
               "devtools", "dplyr", "RColorBrewer", "ggrepel", "ComplexUpset")

# install aamisc package for MDS and Volcano plots if not already done
# Uncomment if needed to install the aamisc package
#pacman::p_load("qvalue", "rain", "limma", "devtools")
#url <- "https://cran.r-project.org/src/contrib/Archive/HarmonicRegression/HarmonicRegression_1.0.tar.gz"
#pkgFile <- "HarmonicRegression_1.0.tar.gz"
#download.file(url = url, destfile = pkgFile)
#install.packages(pkgs=pkgFile, type="source", repos=NULL)
#file.remove(pkgFile)
#pacman::p_load_gh("altintasali/aamisc")

# Color scheme and shapes for figures
RA_colors <- c("#003049", "#0073af")
LA_colors <- c("#d62828", "#e77d7d")
RV_colors <- c("#ff7f00", "#ffb266")
LV_colors <- c("#fcbf49", "#fee2ad")
# Shape 16 (circle) for Control and Shape 17 (triangle) for AF
condition_shapes <- c("Control" = 16, "AF" = 17)
chamber_colors <- c("RA" = "#003049", "LA" = "#d62828", "RV" = "#ff7f00", "LV" = "#fcbf49")
chamber_fill_colors <- c("RA" = "#0073af", "LA" = "#e77d7d", "RV" = "#ffb266", "LV" = "#fee2ad")


```

# Read data

## Count matrix and metadata
```{r read_data}
# Load count data from a local TSV file and metadata through Excel
count_file <- "../../../data/count/gene-expression-all-reverse-stranded-countReadPairs.tsv"
count <- readr::read_delim(count_file)
meta_file <- "../../../data/metadata/meta.xlsx" 
meta <- readxl::read_excel(meta_file)

# Gene annotation
geneinfo_file <- "../../../data/gene_annotation/horse_gene_annotation.tsv.gz"
geneinfo <- fread(geneinfo_file)
colnames(geneinfo)

# Rename columns
setnames(geneinfo, new = c("ENSEMBL", "ENSEMBLv", "Description_detailed", 
                           "Chr", "Start", "End", "Strand", "GENENAME", "ENTREZID", "Description"))
# Remove unnecessary columns
geneinfo <- geneinfo[, c("ENSEMBLv", "Description_detailed") := NULL]

# Remove duplicate entries based on ENSEMBL ID
geneinfo <- geneinfo[!duplicated(ENSEMBL), ]

# Convert annotation data to data frame and set rownames to ENSEMBL IDs
annot <- merge(x = count[,c("Geneid", "Length")], 
               y = geneinfo, 
               by.x = "Geneid",
               by.y = "ENSEMBL", 
               all.x = TRUE, 
               all.y = FALSE)
setnames(annot, old = "Geneid", new = "ENSEMBL")
annot <- data.frame(annot)
rownames(annot) <- annot$ENSEMBL

fwrite(x = annot, 
       file = "../../../data/gene_annotation/horse_gene_annotation_filtered.tsv.gz", 
       sep = "\t")

# Clean metadata and ensure it matches the count data
meta <- meta %>% remove_rownames %>% column_to_rownames(var="Sample_ID")
head(meta)

count <- count[, -c(2:6)]

# Set "Geneid" as rownames and remove from the data frame for easy manipulation
count <- count %>% remove_rownames %>% column_to_rownames(var="Geneid")
head(count)


```


# `edgeR` differential expression analysis for multilevel designs
## Read count matrix

```{r edgeR_read, warning=FALSE}
# Remove version numbers from gene names if present (e.g., "Gene.1" -> "Gene").
rownames(count) <- stringr::str_split_fixed(string = rownames(count), 
                                        pattern = '[.]',
                                        n = 2)[,1]

# Reorder metadata and annotation tables to match the column order of the count matrix.
column_order <- names(count)
meta_reordered <- meta[column_order, , drop = FALSE]

annot_order <- rownames(count)
annot_reordered <- annot[annot_order, ]

# Create a DGEList object for `edgeR` analysis.
# The DGEList object contains the count matrix, gene annotation, and sample metadata.
d <- DGEList(counts = count, genes = annot_reordered, samples = meta_reordered)
```


## Filtering
```{r edgeR_filter, warning=FALSE}
# Define the design matrix based on the experimental conditions
# This design matrix will help retain genes that are expressed in at least one condition
design <- model.matrix(~0 + Condition , d$samples)
colnames(design) <- gsub("Condition", "", colnames(design))
design
keep <- filterByExpr(d, design = design)
table(keep)

# Create a new DGEList object with only the retained genes
y <- d[keep, ,keep.lib.sizes=FALSE]

# Summary of the filtered DGEList object
cat("Number of genes & samples:", dim(y)[1], "&", dim(y)[2], "\n")
```

## Normalization
```{r edgeR_normalize, warning=FALSE}
# calculate normalization factors (TMM normalization of library size using edgeR)
y <- calcNormFactors(y)

# create design matrix
design <- model.matrix(~0 + Condition , y$samples)
colnames(design) <- gsub("Condition", "", colnames(design))
design

# estimate dispersions
y <- estimateDisp(y, design)

# normalized expression levels (log2-transformed version of TMM-normalized CPM-values)
CPM <- cpm(y)
logCPM <- cpm(y, log = TRUE)

# Save the logCPM matrix to a CSV file
# write.csv(logCPM, "../output/Log2_TMM_normalized_CPM_values.csv", row.names = TRUE)


```

### Count distributions
```{r plot_distributions}
# Transform the logCPM matrix into long format for ggplot2 visualization.
# This will allow each gene/sample pair to be plotted individually.
logCPM_melted <- data.table::melt(logCPM)

# Rename columns for better readability and consistency.
# "Var1" and "Var2" refer to the dimensions of the matrix and are renamed to "Gene" and "Sample".
setnames(x = logCPM_melted, 
         old = c("Var1", "Var2", "value"),
         new = c("Gene", "Sample", "logCPM"))

# Add experimental conditions like "Group" or "Region" to each sample for coloring in the plots.
logCPM_melted <- data.table::merge.data.table(x = logCPM_melted, 
                                              y = data.table(Sample = rownames(meta), meta), 
                                              by = "Sample")

# Color the boxes based on the experimental condition to highlight potential differences.
ggplot(logCPM_melted, aes(x = Sample, y = logCPM)) + 
  geom_boxplot(aes(color = Condition)) +   
  theme_bw() +                             
  theme(axis.text.x = element_text(angle = 90, vjust = 0.5, hjust = 1)) +  
  ggtitle("logCPM boxplots")             

# Create density plots of logCPM values for each sample.
# This visualization helps to assess the distribution of expression values and identify batch effects.
ggplot(logCPM_melted, aes(x = logCPM)) + 
  geom_density(aes(group = Sample, color = Condition)) +  
  theme_bw() +                                           
  ggtitle("logCPM density distributions")                
```

## Dimension reduction
### PCA Plot All horses (un-corrected)
```{r PCA, fig.height=10, fig.width=10}
# Define custom colors and shapes for PCA plots
condition_shapes <- c("Control" = 16, "AF" = 17)  # Circle for Control, Triangle for AF

# Perform PCA on the logCPM values
pca <- prcomp(t(logCPM))  # Transpose logCPM so that samples are in rows and genes are in columns

# Extract PCA scores
pca_scores <- as.data.frame(pca$x)  # PCA scores
pca_scores$Sample <- rownames(pca_scores)  # Add sample names

# Merge with metadata for plotting
pca_scores <- merge(pca_scores, meta, by.x = "Sample", by.y = "row.names")

# Define dimensions for plotting
dims <- list(p1 = c("PC1", "PC2"), p2 = c("PC1", "PC3"), p3 = c("PC1", "PC4"))
pca_plot <- list()

# Without labels
for (i in seq_along(dims)){
  pca_plot[[i]] <- ggplot(pca_scores, aes_string(x = dims[[i]][1], y = dims[[i]][2], color = "Region", shape = "Group")) +
    geom_point(size = 4, alpha = 0.8) +
    scale_color_manual(values = chamber_colors) +
    scale_shape_manual(values = condition_shapes) +
    theme_minimal() +
    labs(title = paste("PCA:", dims[[i]][1], "vs", dims[[i]][2]), x = dims[[i]][1], y = dims[[i]][2]) +
    theme(legend.position = "right")
}

# Combine plots using patchwork
patchwork::wrap_plots(pca_plot, ncol = 2) + patchwork::plot_layout(guides = 'collect')

# With labels
for (i in seq_along(dims)){
  pca_plot[[i]] <- ggplot(pca_scores, aes_string(x = dims[[i]][1], y = dims[[i]][2], color = "Region", shape = "Group", label = "Horse")) +
    geom_point(size = 4, alpha = 0.8) +
    geom_text(vjust = -0.5, size = 3) +  # Add labels for each sample
    scale_color_manual(values = chamber_colors) +
    scale_shape_manual(values = condition_shapes) +
    theme_minimal() +
    labs(title = paste("PCA:", dims[[i]][1], "vs", dims[[i]][2]), x = dims[[i]][1], y = dims[[i]][2]) +
    theme(legend.position = "right")
}

# Combine labeled plots using patchwork
patchwork::wrap_plots(pca_plot, ncol = 2) + patchwork::plot_layout(guides = 'collect')

### Extract top loadings
# Define a function to extract the top contributing genes for each component
get_top_loadings <- function(pca_obj, pc_num, top_n = 10) {
  # Extract loadings for the specified component
  loadings <- pca_obj$rotation[, pc_num]
  
  # Get the top genes with highest absolute loadings
  top_genes <- sort(abs(loadings), decreasing = TRUE)[1:top_n]
  
  # Extract gene names and actual loading values for the top genes
  top_loadings <- data.frame(
    Gene = names(top_genes),         # Gene names
    Loading = loadings[names(top_genes)],  # Loading values
    PC = paste0("PC", pc_num)        # Principal Component identifier
  )
  
  return(top_loadings)
}
### Verdict: Batch effect relating to each individual horse contributing four samples

```

### PCA Plot All samples - Batch effect removed - Figure 2
```{r PCA_Batch_Corrected, fig.height=10, fig.width=10}
# Remove batch effect using `removeBatchEffect` before PCA
logCPM_batchRemoved <- removeBatchEffect(logCPM, batch = y$samples$Horse, design = design)

# Perform PCA on batch-corrected logCPM values
pca_batchRemoved <- prcomp(t(logCPM_batchRemoved))

# Extract PCA scores and add sample information
pca_scores <- as.data.frame(pca_batchRemoved$x)  
pca_scores$Sample <- rownames(pca_scores)  

# Merge PCA scores with metadata
pca_scores <- merge(pca_scores, meta, by.x = "Sample", by.y = "row.names")

# Define dimensions for PCA plotting
dims <- list(p1 = c("PC1", "PC2"), p2 = c("PC2", "PC3"), p3 = c("PC1", "PC4"))
pca_plot <- list()

# Create PCA plots without labels
for (i in seq_along(dims)){
  pca_plot[[i]] <- ggplot(pca_scores, aes_string(x = dims[[i]][1], y = dims[[i]][2], color = "Region", shape = "Group")) +
    geom_point(size = 4, alpha = 0.8) +
    scale_color_manual(values = chamber_colors) +
    scale_shape_manual(values = condition_shapes) +
    theme_minimal() +
    labs(title = paste("PCA:", dims[[i]][1], "vs", dims[[i]][2]), x = dims[[i]][1], y = dims[[i]][2]) +
    theme(legend.position = "right")
}

# Combine PCA plots using patchwork
patchwork::wrap_plots(pca_plot, ncol = 2) + patchwork::plot_layout(guides = 'collect')

# Create PCA plots with labels for each sample
for (i in seq_along(dims)){
  pca_plot[[i]] <- ggplot(pca_scores, aes_string(x = dims[[i]][1], y = dims[[i]][2], color = "Region", shape = "Group", label = "Horse")) +
    geom_point(size = 4, alpha = 0.8) +
    geom_text(vjust = -0.5, size = 3) +  # Add labels for each sample
    scale_color_manual(values = chamber_colors) +
    scale_shape_manual(values = condition_shapes) +
    theme_minimal() +
    labs(title = paste("PCA:", dims[[i]][1], "vs", dims[[i]][2]), x = dims[[i]][1], y = dims[[i]][2]) +
    theme(legend.position = "right")
}

# Combine labeled PCA plots using patchwork
patchwork::wrap_plots(pca_plot, ncol = 2) + patchwork::plot_layout(guides = 'collect')

# Find top loadings
## Function to get all loadings for specified principal components and save to one file
extract_all_loadings <- function(pca_obj, pcs = 1:6, annot) {
  # Extract all loadings for the specified principal components
  all_loadings <- pca_obj$rotation[, pcs, drop = FALSE]
  
  # Convert to a data frame for easy manipulation
  loadings_df <- as.data.frame(all_loadings)
  
  # Add ENSEMBL IDs as a column
  loadings_df$ENSEMBL <- rownames(loadings_df)
  
  # Merge with annotation to include gene names
  loadings_df <- merge(loadings_df, annot[, c("ENSEMBL", "GENENAME")], by = "ENSEMBL", all.x = TRUE)
  
  # Rearrange columns to have GENENAME and ENSEMBL first for clarity
  loadings_df <- loadings_df[, c("GENENAME", "ENSEMBL", colnames(loadings_df)[1:(ncol(loadings_df) - 2)])]
  
  return(loadings_df)
}

# Extract loadings for PC1 to PC6
all_loadings_df <- extract_all_loadings(pca_obj = pca_batchRemoved, pcs = 1:6, annot = annot_reordered)

# Save the combined loadings to a single CSV file
write.csv(all_loadings_df, file = "../output/All_Loadings_PC1_to_PC6.csv", row.names = FALSE)

# Preview the combined loadings
head(all_loadings_df)

# Source the helper functions
source("pca_helpers_all_samples.R")

# Function to extract loadings based on highlighted genes
extract_and_map_loadings <- function(pca_obj, gene_list, annot) {
  ensembl_ids <- annot$ENSEMBL[match(gene_list, annot$GENENAME)]
  ensembl_ids <- ensembl_ids[!is.na(ensembl_ids)]
  valid_ids <- intersect(ensembl_ids, rownames(pca_obj$rotation))

  if (length(valid_ids) == 0) {
    warning("No matching ENSEMBL IDs found in PCA loadings.")
    return(NULL)
  }

  loadings_df <- as.data.frame(pca_obj$rotation[valid_ids, , drop = FALSE])
  loadings_df$ENSEMBL <- rownames(loadings_df)
  loadings_df <- merge(loadings_df, annot[, c("ENSEMBL", "GENENAME")], by = "ENSEMBL", all.x = TRUE)
  
  return(loadings_df)
}

### FOR PC1 & PC2

# Highlighted genes for PCA loadings extraction
highlighted_genes <- c("SCD5", "PITX2", "KCNK3", "BMP10", "KCVN2", "IRX2", "NR4A3", "MYH7")
specific_loadings_df <- extract_and_map_loadings(pca_batchRemoved, highlighted_genes, annot_reordered)
pca_batchRemoved <- prcomp(t(logCPM_batchRemoved))
pca_scores <- as.data.frame(pca_batchRemoved$x)  # PCA scores
pca_scores$Sample <- rownames(pca_scores)  # Add sample names
pca_scores <- merge(pca_scores, meta, by.x = "Sample", by.y = "row.names")

# Generate PCA plots without and with labels
dimensions <- list(c("PC1", "PC2"))

# Generate and display PCA plots for all samples (both healthy and disease groups)
generate_and_save_pca_plots(
  pca_df = pca_scores,
  loadings_df = specific_loadings_df,
  dimensions = dimensions,
  colors = chamber_colors,
  shapes = condition_shapes,  # Include shapes for control/disease
  add_ellipse = FALSE,  # Optionally toggle to TRUE if ellipses are needed
  shape_var = "Group"  # Use Group as the shape variable
)

### FOR PC2 & PC3, PC1 & PC4

# Highlighted genes for PCA loadings extraction
highlighted_genes <- c("NPPA", "KCNJ5", "KCNK3", "NPPB", "MYL7", "KCNV2", "PITX2", "BMP10", 
                       "IRX4", "IRX5", "IRX3", "KCNJ2", "SCD5")
specific_loadings_df <- extract_and_map_loadings(pca_batchRemoved, highlighted_genes, annot_reordered)
print(specific_loadings_df)
pca_batchRemoved <- prcomp(t(logCPM_batchRemoved))
pca_scores <- as.data.frame(pca_batchRemoved$x)  # PCA scores
pca_scores$Sample <- rownames(pca_scores)  # Add sample names
pca_scores <- merge(pca_scores, meta, by.x = "Sample", by.y = "row.names")

dimensions <- list(c("PC1", "PC3"), c("PC1", "PC4"))
generate_and_save_pca_plots(
  pca_df = pca_scores,
  loadings_df = specific_loadings_df,
  dimensions = dimensions,
  colors = chamber_colors,
  shapes = condition_shapes,  # Include shapes for control/disease
  add_ellipse = FALSE,  # Optionally toggle to TRUE if ellipses are needed
  shape_var = "Group"  # Use Group as the shape variable
)

### Extract variance explained by each component
# Extract the standard deviation of each principal component
pca_sdev <- pca_batchRemoved$sdev

# Calculate the variance explained by each component
variance_explained <- (pca_sdev^2) / sum(pca_sdev^2)

# Convert to a data frame for easier viewing
variance_explained_df <- data.frame(
  Principal_Component = paste0("PC", 1:length(variance_explained)),
  Variance_Explained = variance_explained,
  Cumulative_Variance = cumsum(variance_explained)
)

# Print variance explained by each component
print(variance_explained_df)

```

### Subset to only healthy horses for figure 1
```{r PCA_Batch_Corrected_Healthy, fig.height=10, fig.width=10}
# Subset the metadata to include only "Control" horses
healthy_meta <- meta[meta$Group == "Control", ]

# Subset the count matrix to include only columns (samples) of healthy horses
healthy_samples <- rownames(healthy_meta)  # Get sample names of healthy horses
logCPM_healthy <- logCPM[, healthy_samples]  # Subset the logCPM matrix

# Use removeBatchEffect with healthy data only
logCPM_healthy_batchRemoved <- removeBatchEffect(logCPM_healthy, 
                                                 batch = y$samples[healthy_samples, ]$Horse, 
                                                 design = design[healthy_samples, ])

#Perform PCA on the healthy horses' logCPM values
pca_healthy <- prcomp(t(logCPM_healthy_batchRemoved))

# Extract PCA scores for healthy horses
pca_scores_healthy <- as.data.frame(pca_healthy$x)  # PCA scores for healthy horses
pca_scores_healthy$Sample <- rownames(pca_scores_healthy)  # Add sample names

# Merge with the healthy subset of metadata for plotting
pca_scores_healthy <- merge(pca_scores_healthy, healthy_meta, by.x = "Sample", by.y = "row.names")

# Define dimensions for PCA plotting 
dims <- list(p1 = c("PC1", "PC2"), p2 = c("PC1", "PC3"), p3 = c("PC2", "PC3"))
pca_plot_healthy <- list()

# Create PCA plots for healthy horses using predefined colors and shapes
for (i in seq_along(dims)){
  pca_plot_healthy[[i]] <- ggplot(pca_scores_healthy, aes_string(x = dims[[i]][1], y = dims[[i]][2], 
                                                                color = "Region", fill = "Region")) +
    geom_point(size = 4, alpha = 0.8, stroke = 1.5) +  # Adjusted to include border color
    scale_color_manual(values = chamber_colors) +      # Outer colors
    scale_fill_manual(values = chamber_fill_colors) +  # Fill colors
    scale_shape_manual(values = condition_shapes) +
    theme_minimal() +
    labs(title = paste("PCA (Healthy Horses):", dims[[i]][1], "vs", dims[[i]][2]), 
         x = dims[[i]][1], y = dims[[i]][2]) +
    theme(legend.position = "right")
}
patchwork::wrap_plots(pca_plot_healthy, ncol = 2) + patchwork::plot_layout(guides = 'collect')

# With labels
for (i in seq_along(dims)){
  pca_plot_healthy[[i]] <- ggplot(pca_scores_healthy, aes_string(x = dims[[i]][1], y = dims[[i]][2], color = "Region", shape = "Group", label = "Horse")) +
    geom_point(size = 4, alpha = 0.8) +
    geom_text(vjust = -0.5, size = 3) +  # Add labels for each sample
    scale_color_manual(values = chamber_colors) +
    scale_shape_manual(values = condition_shapes) +
    theme_minimal() +
    labs(title = paste("PCA:", dims[[i]][1], "vs", dims[[i]][2]), x = dims[[i]][1], y = dims[[i]][2]) +
    theme(legend.position = "right")
}

# Combine PCA plots for healthy horses using patchwork
patchwork::wrap_plots(pca_plot_healthy, ncol = 2) + patchwork::plot_layout(guides = 'collect')

```

#### Add Loadings
```{r PCA_with_loadings, fig.height=10, fig.width=10}
# Source the helper functions
source("pca_helpers.R")

#Extract loadings for plot
# Unified function to extract specific loadings with gene names
extract_and_map_loadings <- function(pca_obj, gene_list, annot) {
  # Map gene symbols to ENSEMBL IDs and remove NAs
  ensembl_ids <- annot$ENSEMBL[match(gene_list, annot$GENENAME)]
  ensembl_ids <- ensembl_ids[!is.na(ensembl_ids)]

  # Extract loadings for the mapped ENSEMBL IDs present in PCA loadings
  valid_ids <- intersect(ensembl_ids, rownames(pca_obj$rotation))
  if (length(valid_ids) == 0) {
    warning("No matching ENSEMBL IDs found in PCA loadings.")
    return(NULL)
  }
  
  # Extract loadings and add gene names from annotation
  loadings_df <- as.data.frame(pca_obj$rotation[valid_ids, , drop = FALSE])
  loadings_df$ENSEMBL <- rownames(loadings_df)
  loadings_df <- merge(loadings_df, annot[, c("ENSEMBL", "GENENAME")], by = "ENSEMBL", all.x = TRUE)
  
  return(loadings_df)
}

# Example gene list for loadings extraction (modify as needed)
highlighted_genes <- c("BMP10", "NPPA", "NPPB", "KCNJ5", "KCNK3", "MYL7", "KCNV2", "PCDH7", "KIF18B")

# Extract loadings with unified function
specific_loadings_df <- extract_and_map_loadings(pca_healthy, highlighted_genes, annot_reordered)

# Check results
print(specific_loadings_df)

# Example dimensions for PCA plotting
dimensions <- list(c("PC1", "PC2"), c("PC1", "PC3"), c("PC2", "PC3"))

# Generate and display the PCA plots with probability ellipses
generate_and_save_pca_plots(
  pca_df = pca_scores_healthy,
  loadings_df = specific_loadings_df,
  dimensions = dimensions,
  colors = chamber_colors,
  add_ellipse = FALSE  # Toggle to TRUE to include ellipses
)


### Extract variance explained by each component
# Extract the standard deviation of each principal component
pca_sdev <- pca_healthy$sdev

# Calculate the variance explained by each component
variance_explained <- (pca_sdev^2) / sum(pca_sdev^2)

# Convert to a data frame for easier viewing
variance_explained_df <- data.frame(
  Principal_Component = paste0("PC", 1:length(variance_explained)),
  Variance_Explained = variance_explained,
  Cumulative_Variance = cumsum(variance_explained)
)

# Print variance explained by each component
print(variance_explained_df)
```


# Differential expression analysis 

## Create contrast for multilevel design
```{r edgeR_contrast}
# create contrasts to test
colnames(design)
con <- makeContrasts(AF_vs_control_RA = RA_AF - RA_Control,
                     AF_vs_control_LA = LA_AF - LA_Control,
                     RA_vs_LA_control = RA_Control - LA_Control,
                     RA_vs_LA_AF = RA_AF - LA_AF,
                     InteractionEffect_Atrial = (RA_AF - RA_Control) - (LA_AF - LA_Control),
                     Avr.Dis.Effect_Atrial = (LA_AF + RA_AF)/2 - (LA_Control + RA_Control)/2,
                     AverageRegionEffect_Atrial = (RA_Control + RA_AF)/2 - (LA_Control + LA_AF)/2,
                     AF_vs_control_RV = RV_AF - RV_Control,
                     AF_vs_control_LV = LV_AF - LV_Control,
                     RV_vs_LV_control = RV_Control - LV_Control,
                     RV_vs_LV_AF = RV_AF - LV_AF,
                     InteractionEffect_Ventricular = (RV_AF - RV_Control) - (LV_AF - LV_Control),
                     Avr.Dis.Effect_Ventricular = (LV_AF + RV_AF)/2 - (LV_Control + RV_Control)/2,
                     AverageRegionEffect_Ventricular = (RV_Control + RV_AF)/2 - (LV_Control + LV_AF)/2,
                     AverageRegionEffect_AV_healthy = (RA_Control + LA_Control)/2 - (RV_Control + LV_Control)/2,
                     levels = design)
con

# Main effects
con_disease <- makeContrasts(AF_vs_control_RA = RA_AF - RA_Control,
                             AF_vs_control_LA = LA_AF - LA_Control,
                             AF_vs_control_RV = RV_AF - RV_Control,
                             AF_vs_control_LV = LV_AF - LV_Control,
                             levels = design)

con_region <- makeContrasts( RA_vs_LA_control = RA_Control - LA_Control,
                            RA_vs_LA_AF = RA_AF - LA_AF,
                            RV_vs_LV_control = RV_Control - LV_Control,
                            RV_vs_LV_AF = RV_AF - LV_AF,
                            levels = design)

```

## Differential analysis
```{r voomLmFit, fig.height=7, fig.width=7}
# V is the result of my linear model fitted using voom transformation
y_raw <- d[keep, ,keep.lib.sizes=FALSE]
v <- voomLmFit(counts = y_raw, 
               design = design, 
               block = as.factor(y_raw$samples$Horse), 
               sample.weights = TRUE, 
               plot = TRUE) 

res <- list() # list for DGE results
for (i in colnames(con)) {
  fit <- contrasts.fit(v, contrasts = con)
  fit <- eBayes(fit, robust = TRUE)
  res[[i]] <- topTable(fit, coef = i, number = Inf)
  res[[i]] <- data.frame(res[[i]], Contrast = i)
  n <- res[[i]] %>% dplyr::filter(adj.P.Val < 0.05) %>% nrow 
  print(paste('number of DE genes for',i, '=', n))
}
res_all <- do.call(rbind, res)

# Create output Excel file
openxlsx::write.xlsx(x = res, file = "../output/dge_results.xlsx", asTable = TRUE)

# Create output TSV file
data.table::fwrite(x = res_all, file = "../output/dge_results.tsv.gz", sep = "\t")
```


## Diagnostics for differential analysis

### p-value histograms
```{r pvalue_histograms, fig.height=10, fig.width=10}
ggplot(res_all, aes(x = P.Value)) + 
  geom_histogram(fill = "lightgray",
                 color = "black",
                 breaks = seq(0, 1, by = 0.05),
                 closed = "right",
                 lwd = 0.2) + 
  facet_wrap(~ Contrast, nrow = 3, scales = "free") + 
  theme_bw()
```

### Volcano plots
```{r volcano_plots, fig.height=10, fig.width=10}
volcano_plots <- list()
for (i in names(res)){
  volcano_plots[[i]] <- ggVolcano(x = res[[i]], 
                                  fdr = 0.05,
                                  fdr.column = "adj.P.Val", 
                                  pvalue.column = "P.Value", 
                                  logFC = 0, 
                                  logFC.column = "logFC", 
                                  text.size = 2) + 
    theme_bw(base_size = 10) + 
    ggtitle(i)
}

# Combine all volcano plots into a single layout with 3 columns.
patchwork::wrap_plots(volcano_plots, ncol = 3)

# Print individual volcano plots for key contrasts.
print(volcano_plots[["RA_vs_LA_control"]])
print(volcano_plots[["RV_vs_LV_control"]])
print(volcano_plots[["AverageRegionEffect_AV_healthy"]])

# Create a combined plot of the key contrasts for easier comparison.
combined_plot <- ((volcano_plots[["RA_vs_LA_control"]]) | (volcano_plots[["AverageRegionEffect_AV_healthy"]]))
print(combined_plot)
```
### Volcano Plot Custom
```{r volcano_plots_custom, fig.height=10, fig.width=10}
regions <- unique(healthy_meta$Region)

# Gene Name Mapping for Volcano Plots in RNA-seq Analysis
# Check if the annotation dataframe has the necessary columns
if (!"ENSEMBL" %in% colnames(annot_reordered) || !"GENENAME" %in% colnames(annot_reordered)) {
  stop("Error: The annotation dataframe must contain both 'ENSEMBL' and 'GENENAME' columns.")
}
# Create a named vector for ENSEMBL to GENENAME mapping
ensembl_to_genename <- setNames(annot_reordered$GENENAME, annot_reordered$ENSEMBL)
# Map ENSEMBL IDs to GeneNames in the res list
for (contrast_name in names(res)) {
  # Ensure the dataframe has ENSEMBL IDs as rownames
  if (!"ENSEMBL" %in% colnames(res[[contrast_name]])) {
    res[[contrast_name]]$ENSEMBL <- rownames(res[[contrast_name]])
  }
  
  # Map GENENAMEs to the dataframe using ENSEMBL IDs
  res[[contrast_name]]$GENENAME <- ensembl_to_genename[res[[contrast_name]]$ENSEMBL]
  
  # Replace NA values in GENENAME with ENSEMBL IDs (to ensure plotting works even if some gene names are missing)
  res[[contrast_name]]$GENENAME[is.na(res[[contrast_name]]$GENENAME)] <- res[[contrast_name]]$ENSEMBL[is.na(res[[contrast_name]]$GENENAME)]
}


# Source the helper function
source("volcano_helpers.R")

# Create lists to store both versions of volcano plots
volcano_plots_no_labels <- list()
volcano_plots_with_labels <- list()

# Iterate over each contrast in `res` and create custom volcano plots with/without labels
for (contrast_name in names(res)) {
  # Ensure the GeneName column is present in the dataframe for labeling
  if (!"GeneName" %in% colnames(res[[contrast_name]])) {
    # Map ENSEMBL to GeneName using the preprocessed mapping vector
    res[[contrast_name]]$GeneName <- sapply(rownames(res[[contrast_name]]), function(x) gsub(".*_", "", x))
  }
  
  # Generate volcano plots with and without labels using the helper function
  volcano_plots <- create_custom_volcano_plot(
    df = res[[contrast_name]],
    logFC_col = "logFC",
    pvalue_col = "P.Value",
    adj_pvalue_col = "adj.P.Val",
    contrast_name = contrast_name,
    fc_cutoff = 1,
    pvalue_cutoff = 0.05,
    save_plot = TRUE, 
    output_path = "../output/",
    show_labels = TRUE  # Always generate both labeled and unlabeled versions
  )
  
  # Store the plots in separate lists
  volcano_plots_no_labels[[contrast_name]] <- volcano_plots$No_Labels
  volcano_plots_with_labels[[contrast_name]] <- volcano_plots$With_Labels
}

# Combine and display the volcano plots without labels
patchwork::wrap_plots(volcano_plots_no_labels, ncol = 3)

# Print individual volcano plots with labels for key contrasts
print(volcano_plots_with_labels[["RA_vs_LA_control"]])
print(volcano_plots_no_labels[["RA_vs_LA_control"]])
print(volcano_plots_with_labels[["RV_vs_LV_control"]])
print(volcano_plots_with_labels[["AverageRegionEffect_AV_healthy"]])
print(volcano_plots_with_labels[["AF_vs_control_LA"]])
print(volcano_plots_with_labels[["AF_vs_control_RA"]])

```


<!-- # Atrial Selectivty based on count -->
<!-- ```{r, Overlap, fig.height=5, fig.width=5} -->
<!-- #  Filter metadata to include only healthy (Control) horses -->
<!-- healthy_meta <- meta %>% filter(Group == "Control") -->

<!-- #  Subset the filtered count matrix (y$counts) to include only columns (samples) from healthy horses -->
<!-- healthy_samples <- rownames(healthy_meta) -->
<!-- healthy_filtered_counts <- y$counts[, healthy_samples, drop = FALSE] -->

<!-- # Set the minimum count threshold for detection and the percentage of samples in a region to consider a gene detected -->
<!-- count_threshold <- 5  # Minimum count to consider a gene detected -->
<!-- detection_percentage <- 0.75  # Minimum percentage of samples in a region for detection -->

<!--   # TODO check with Ali if these thresholds are relevant  -->

<!-- #  Filter count data for detected genes in each region using the threshold and detection percentage -->
<!-- detected_genes_list <- list() -->

<!-- for (region in regions) { -->
<!--   region_samples <- rownames(healthy_meta[healthy_meta$Region == region, ]) -->
<!--   region_filtered_counts <- healthy_filtered_counts[, region_samples, drop = FALSE] -->

<!--   # Identify genes with counts greater than or equal to the threshold in at least 'detection_percentage' of samples -->
<!--   detected_genes <- rownames(region_filtered_counts[ -->
<!--     rowMeans(region_filtered_counts >= count_threshold) >= detection_percentage,  -->
<!--   ]) -->

<!--   # Debug: Display number of detected genes in each region -->
<!--   cat("Number of detected genes in", region, ":", length(detected_genes), "\n") -->

<!--   detected_genes_list[[region]] <- detected_genes -->
<!-- } -->

<!-- # Create a binary matrix for gene presence/absence in each region based on the detection criteria -->
<!-- all_detected_genes <- unique(unlist(detected_genes_list)) -->

<!-- # Create a binary matrix with rows as genes and columns as regions -->
<!-- gene_presence_matrix <- data.frame( -->
<!--   Gene = all_detected_genes, -->
<!--   LA = as.numeric(all_detected_genes %in% detected_genes_list[["LA"]]), -->
<!--   RA = as.numeric(all_detected_genes %in% detected_genes_list[["RA"]]), -->
<!--   LV = as.numeric(all_detected_genes %in% detected_genes_list[["LV"]]), -->
<!--   RV = as.numeric(all_detected_genes %in% detected_genes_list[["RV"]]) -->
<!-- ) -->

<!-- # Create the UpSet plot using ComplexUpset -->
<!-- upset_data <- gene_presence_matrix %>% -->
<!--   mutate(across(-Gene, as.logical))  # Convert columns to logical values -->
<!-- # Define set names for easy reference -->
<!-- set_names <- c("LA", "RA", "LV", "RV") -->

<!-- # Create the UpSet plot to show overlap of detected genes across regions -->
<!-- upset_plot <- ComplexUpset::upset( -->
<!--   upset_data, set_names,  -->
<!--   name = "Region", -->
<!--   intersections = list( -->
<!--     c("LA"), c("RA"), c("LV"), c("RV"),  # Individual sets -->
<!--     c("LA", "RA"), c("LV", "RV"),         # Atrial and ventricular combinations -->
<!--     c("LA", "LV"), c("RA", "RV"),         # Left and right side combinations -->
<!--     c("LA", "RA", "LV", "RV"),             # Genes present in all four regions -->
<!--     c("LA", "RV"), c("RA","LV")          # Genes "asymmetrically" present in right/left atria/ventricle -->
<!--   ), -->
<!--   min_size = 10,  # Minimum size of intersections to be displayed -->
<!--   width_ratio = 0.2, height_ratio = 0.4,  # Adjust ratios for better visualization -->
<!--   themes = upset_default_themes(panel.grid.major = element_blank()) -->
<!-- ) -->

<!-- # Add annotations and customize the plot -->
<!-- upset_plot <- upset_plot +  -->
<!--   ggtitle("Overlap of Robustly Detected Genes in Healthy Horse Heart Regions") +  -->
<!--   theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold")) -->

<!-- # Display ans save the plot -->
<!-- print(upset_plot) -->
<!-- ggsave(filename = "../output/Detected_Genes_Intersections_UpSet.png",  # Specify file path and name -->
<!--   plot = upset_plot,                                              # Use the created upset_plot object -->
<!--   width = 6, height = 4,                                         # Set desired width and height in inches -->
<!--   dpi = 600                                                       # Set the resolution (DPI) for high quality -->
<!-- ) -->


<!-- ###### EXPORT ######### -->
<!-- #  Map ENSEMBL IDs to Gene Names -->
<!-- upset_data_annot <- merge(upset_data, annot_reordered[, c("ENSEMBL", "GENENAME")],  -->
<!--                           by.x = "Gene", by.y = "ENSEMBL", all.x = TRUE) -->

<!-- # Replace NA values in `GENENAME` with ENSEMBL IDs to ensure every gene has a name -->
<!-- upset_data_annot$GENENAME[is.na(upset_data_annot$GENENAME)] <- upset_data_annot$Gene[is.na(upset_data_annot$GENENAME)] -->

<!-- # Create a named list to store gene lists for each intersection -->
<!-- gene_lists <- list() -->

<!-- #  Define and extract genes for each intersection of interest -->
<!-- intersections <- list( -->
<!--   "LA_Only" = list(include = c("LA"), exclude = c("RA", "LV", "RV")), -->
<!--   "RA_Only" = list(include = c("RA"), exclude = c("LA", "LV", "RV")), -->
<!--   "LV_Only" = list(include = c("LV"), exclude = c("LA", "RA", "RV")), -->
<!--   "RV_Only" = list(include = c("RV"), exclude = c("LA", "RA", "LV")), -->
<!--   "Atrial_Specific" = list(include = c("LA", "RA"), exclude = c("LV", "RV")), -->
<!--   "Ventricular_Specific" = list(include = c("LV", "RV"), exclude = c("LA", "RA")), -->
<!--   "Left_Side_Specific" = list(include = c("LA", "LV"), exclude = c("RA", "RV")), -->
<!--   "Right_Side_Specific" = list(include = c("RA", "RV"), exclude = c("LA", "LV")), -->
<!--   "All_Four_Regions" = list(include = c("LA", "RA", "LV", "RV"), exclude = c()), -->
<!--   "LA_RV" = list(include = c("LA", "RV"), exclude = c("RA", "LV")), -->
<!--   "RA_LV" = list(include = c("RA", "LV"), exclude = c("LA", "RV")) -->
<!-- ) -->

<!-- # Extract genes for each intersection -->
<!-- for (name in names(intersections)) { -->
<!--   include_regions <- intersections[[name]]$include -->
<!--   exclude_regions <- intersections[[name]]$exclude -->

<!--   # Identify genes present in the included regions and absent in the excluded regions -->
<!--   genes_in_intersection <- upset_data_annot %>% -->
<!--     filter(across(all_of(include_regions), ~ . == TRUE)) %>%   # Genes present in included regions -->
<!--     filter(across(all_of(exclude_regions), ~ . == FALSE)) %>%  # Genes absent in excluded regions -->
<!--     select(Gene, GENENAME)  # Select only ENSEMBL ID and gene name columns -->

<!--   # Store the resulting gene list in the `gene_lists` list -->
<!--   gene_lists[[name]] <- genes_in_intersection -->
<!-- } -->

<!-- #  Write each gene list to a separate sheet in an Excel file -->
<!-- output_file <- "../output/Detected_Genes_Intersections.xlsx" -->
<!-- wb <- createWorkbook() -->

<!-- for (sheet_name in names(gene_lists)) { -->
<!--   addWorksheet(wb, sheet_name) -->
<!--   writeData(wb, sheet = sheet_name, x = gene_lists[[sheet_name]]) -->
<!-- } -->

<!-- # Save the Excel file with all intersection lists -->
<!-- saveWorkbook(wb, output_file, overwrite = TRUE) -->

<!-- # Confirmation message -->
<!-- cat("Intersection lists successfully saved to:", output_file, "\n") -->

<!-- ``` -->

# Atrial Selectivity based on RPKM
```{r}
# Calculate RPKM for all genes
total_reads <- colSums(y$counts)  # Total reads per sample
gene_lengths <- annot_reordered$Length  # Gene lengths in base pairs

# Calculate RPKM: (Counts * 1e9) / (Gene Length * Total Reads)
rpkm_matrix <- sweep(y$counts, 2, total_reads, "/") * 1e9
rpkm_matrix <- sweep(rpkm_matrix, 1, gene_lengths, "/")

# Filter metadata to include only healthy (Control) horses
healthy_meta <- meta %>% filter(Group == "Control")

# Subset the filtered RPKM matrix to include only columns (samples) from healthy horses
healthy_samples <- rownames(healthy_meta)
healthy_rpkm_matrix <- rpkm_matrix[, healthy_samples, drop = FALSE]

# Set the minimum RPKM threshold for detection and the percentage of samples in a region to consider a gene detected
rpkm_threshold <- 1  # Minimum RPKM to consider a gene detected
detection_percentage <- 0.5  # Minimum percentage of samples in a region for detection

# Filter RPKM data for detected genes in each region using the threshold and detection percentage
detected_genes_list <- list()

for (region in regions) {
  region_samples <- rownames(healthy_meta[healthy_meta$Region == region, ])
  region_filtered_rpkm <- healthy_rpkm_matrix[, region_samples, drop = FALSE]

  # Identify genes with RPKM greater than or equal to the threshold in at least 'detection_percentage' of samples
  detected_genes <- rownames(region_filtered_rpkm[
    rowMeans(region_filtered_rpkm >= rpkm_threshold) >= detection_percentage, 
  ])
  
  # Debug: Display number of detected genes in each region
  cat("Number of detected genes in", region, ":", length(detected_genes), "\n")

  detected_genes_list[[region]] <- detected_genes
}

# Create a binary matrix for gene presence/absence in each region based on the detection criteria
all_detected_genes <- unique(unlist(detected_genes_list))

# Create a binary matrix with rows as genes and columns as regions
gene_presence_matrix <- data.frame(
  Gene = all_detected_genes,
  LA = as.numeric(all_detected_genes %in% detected_genes_list[["LA"]]),
  RA = as.numeric(all_detected_genes %in% detected_genes_list[["RA"]]),
  LV = as.numeric(all_detected_genes %in% detected_genes_list[["LV"]]),
  RV = as.numeric(all_detected_genes %in% detected_genes_list[["RV"]])
)

# Create the UpSet plot using ComplexUpset
upset_data <- gene_presence_matrix %>%
  mutate(across(-Gene, as.logical))  # Convert columns to logical values

# Define set names for easy reference
set_names <- c("LA", "RA", "LV", "RV")

# Create the UpSet plot to show overlap of detected genes across regions
upset_plot <- ComplexUpset::upset(
  upset_data, set_names, 
  name = "Region",
  intersections = list(
    c("LA"), c("RA"), c("LV"), c("RV"),  # Individual sets
    c("LA", "RA"), c("LV", "RV"),         # Atrial and ventricular combinations
    c("LA", "LV"), c("RA", "RV"),         # Left and right side combinations
    c("LA", "RA", "LV", "RV"),             # Genes present in all four regions
    c("LA", "RV"), c("RA","LV")          # Genes "asymmetrically" present in right/left atria/ventricle
  ),
  min_size = 10,  # Minimum size of intersections to be displayed
  width_ratio = 0.2, height_ratio = 0.4,  # Adjust ratios for better visualization
  themes = upset_default_themes(panel.grid.major = element_blank())
)

# Add annotations and customize the plot
upset_plot <- upset_plot + 
  ggtitle("Overlap of Robustly Detected Genes in Healthy Horse Heart Regions") + 
  theme(plot.title = element_text(hjust = 0.5, size = 16, face = "bold"))

# Display and save the plot
print(upset_plot)
ggsave(
  filename = "../output/Detected_Genes_Intersections_UpSet_RPKM.png",  # Specify file path and name
  plot = upset_plot,                                              # Use the created upset_plot object
  width = 6, height = 4,                                         # Set desired width and height in inches
  dpi = 600                                                       # Set the resolution (DPI) for high quality
)

###### EXPORT #########
# Map ENSEMBL IDs to Gene Names
upset_data_annot <- merge(upset_data, annot_reordered[, c("ENSEMBL", "GENENAME")], 
                          by.x = "Gene", by.y = "ENSEMBL", all.x = TRUE)

# Replace NA values in `GENENAME` with ENSEMBL IDs to ensure every gene has a name
upset_data_annot$GENENAME[is.na(upset_data_annot$GENENAME)] <- upset_data_annot$Gene[is.na(upset_data_annot$GENENAME)]

# Create a named list to store gene lists for each intersection
gene_lists <- list()

# Define and extract genes for each intersection of interest
intersections <- list(
  "LA_Only" = list(include = c("LA"), exclude = c("RA", "LV", "RV")),
  "RA_Only" = list(include = c("RA"), exclude = c("LA", "LV", "RV")),
  "LV_Only" = list(include = c("LV"), exclude = c("LA", "RA", "RV")),
  "RV_Only" = list(include = c("RV"), exclude = c("LA", "RA", "LV")),
  "Atrial_Specific" = list(include = c("LA", "RA"), exclude = c("LV", "RV")),
  "Ventricular_Specific" = list(include = c("LV", "RV"), exclude = c("LA", "RA")),
  "Left_Side_Specific" = list(include = c("LA", "LV"), exclude = c("RA", "RV")),
  "Right_Side_Specific" = list(include = c("RA", "RV"), exclude = c("LA", "LV")),
  "All_Four_Regions" = list(include = c("LA", "RA", "LV", "RV"), exclude = c()),
  "LA_RV" = list(include = c("LA", "RV"), exclude = c("RA", "LV")),
  "RA_LV" = list(include = c("RA", "LV"), exclude = c("LA", "RV"))
)

# Extract genes for each intersection
for (name in names(intersections)) {
  include_regions <- intersections[[name]]$include
  exclude_regions <- intersections[[name]]$exclude
  
  # Identify genes present in the included regions and absent in the excluded regions
  genes_in_intersection <- upset_data_annot %>%
    filter(across(all_of(include_regions), ~ . == TRUE)) %>%   # Genes present in included regions
    filter(across(all_of(exclude_regions), ~ . == FALSE)) %>%  # Genes absent in excluded regions
    select(Gene, GENENAME)  # Select only ENSEMBL ID and gene name columns
  
  # Store the resulting gene list in the `gene_lists` list
  gene_lists[[name]] <- genes_in_intersection
}

# Write each gene list to a separate sheet in an Excel file
output_file <- "../output/Detected_Genes_Intersections_RPKM.xlsx"
wb <- createWorkbook()

for (sheet_name in names(gene_lists)) {
  addWorksheet(wb, sheet_name)
  writeData(wb, sheet = sheet_name, x = gene_lists[[sheet_name]])
}

# Save the Excel file with all intersection lists
saveWorkbook(wb, output_file, overwrite = TRUE)

# Confirmation message
cat("Intersection lists successfully saved to:", output_file, "\n")

```


# Session Info
```{r session_info}
sessionInfo()
```