---
title: "Ion_channel"
author: "Simon Haugaard"
date: "`r Sys.Date()`"
output: 
  html_document:
    toc: true
    toc_float: true
    toc_collapsed: true
    toc_depth: 3
    number_sections: true
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

# Required R libraries
```{r library}
if (!require("pacman")) install.packages("pacman")
pacman::p_load("edgeR", "readr", "readxl", "biomaRt", "magrittr", "tibble", "stringr", 
               "ggplot2", "data.table", "patchwork", "openxlsx", "aamisc", "limma", 
               "devtools", "dplyr", "RColorBrewer", "ggrepel", "ComplexUpset")

# install aamisc package for MDS and Volcano plots if not already done
# Uncomment if needed to install the aamisc package
#pacman::p_load("qvalue", "rain", "limma", "devtools")
#url <- "https://cran.r-project.org/src/contrib/Archive/HarmonicRegression/HarmonicRegression_1.0.tar.gz"
#pkgFile <- "HarmonicRegression_1.0.tar.gz"
#download.file(url = url, destfile = pkgFile)
#install.packages(pkgs=pkgFile, type="source", repos=NULL)
#file.remove(pkgFile)
#pacman::p_load_gh("altintasali/aamisc")

# Color scheme and shapes for figures
RA_colors <- c("#003049", "#0073af")
LA_colors <- c("#d62828", "#e77d7d")
RV_colors <- c("#ff7f00", "#ffb266")
LV_colors <- c("#fcbf49", "#fee2ad")
# Shape 16 (circle) for Control and Shape 17 (triangle) for AF
condition_shapes <- c("Control" = 16, "AF" = 17)
chamber_colors <- c("RA" = "#003049", "LA" = "#d62828", "RV" = "#ff7f00", "LV" = "#fcbf49")
chamber_fill_colors <- c("RA" = "#0073af", "LA" = "#e77d7d", "RV" = "#ffb266", "LV" = "#fee2ad")


```

# Read data

## Count matrix and metadata
```{r read_data}
# Load count data from a local TSV file and metadata through Excel
meta_file <- "../../../data/metadata/meta.xlsx" 
meta <- readxl::read_excel(meta_file)

# Gene annotation
geneinfo_file <- "../../../data/gene_annotation/horse_gene_annotation.tsv.gz"
geneinfo <- fread(geneinfo_file)
colnames(geneinfo)

# Load DGE
dge_file <- "../../01_dge/output/dge_results.tsv.gz" 
dge <- fread(dge_file)

# Load logCPM 
logCPM_file <- "../../01_dge/output/Log2_TMM_normalized_CPM_values.csv"
logCPM <- fread(logCPM_file)

# Rename columns
setnames(geneinfo, new = c("ENSEMBL", "ENSEMBLv", "Description_detailed", 
                           "Chr", "Start", "End", "Strand", "GENENAME", "ENTREZID", "Description"))
# Remove unnecessary columns
geneinfo <- geneinfo[, c("ENSEMBLv", "Description_detailed") := NULL]

# Remove duplicate entries based on ENSEMBL ID
geneinfo <- geneinfo[!duplicated(ENSEMBL), ]

# Convert annotation data to data frame and set rownames to ENSEMBL IDs
annot <- merge(x = count[,c("Geneid", "Length")], 
               y = geneinfo, 
               by.x = "Geneid",
               by.y = "ENSEMBL", 
               all.x = TRUE, 
               all.y = FALSE)
setnames(annot, old = "Geneid", new = "ENSEMBL")
annot <- data.frame(annot)
rownames(annot) <- annot$ENSEMBL

fwrite(x = annot, 
       file = "../../../data/gene_annotation/horse_gene_annotation_filtered.tsv.gz", 
       sep = "\t")

# Clean metadata and ensure it matches the count data
meta <- meta %>% remove_rownames %>% column_to_rownames(var="Sample_ID")
head(meta)

count <- count[, -c(2:6)]

# Set "Geneid" as rownames and remove from the data frame for easy manipulation
count <- count %>% remove_rownames %>% column_to_rownames(var="Geneid")
head(count)
```

#Ionchannel differences in logCPM
##Using GO Annotation
```{r ionchannels}
#Get GO annotation 
go_file <- "../../../data/gene_annotation/horse_GO.tsv.gz"
go <- fread(go_file)
go <- go[, c("Gene stable ID", "GO term accession", "GO term name", "GO domain")]
setnames(go, new = c("ENSEMBL", "GOID", "Description", "GOdomain"))
go <- go[GOdomain != "", ]

#Filter for GO:0034702  ("ion channel complex")
go_0034702 <- go[GOID == "GO:0034702", "ENSEMBL"]
ionchannel_logCPM <- as.data.frame(logCPM[row.names(logCPM) %in% go_0034702[["ENSEMBL"]], ])
ionchannel_logCPM_with_genes <- merge(ionchannel_logCPM, annot, by.x = "row.names", by.y = "ENSEMBL", all.x = TRUE)
ionchannel_melted <- data.table::melt(ionchannel_logCPM_with_genes)
setnames(x = ionchannel_melted, 
         old = c("GENENAME", "variable", "value"),
         new = c("Gene", "Sample", "logCPM"))
ionchannel_melted <- data.table::merge.data.table(x = ionchannel_melted, 
                                              y = data.table(Sample = rownames(meta), meta), 
                                              by = "Sample")
#Plot 
library(dplyr)
mean_expression <- ionchannel_melted %>% group_by(Gene, Condition, Region, Group) %>%
  summarize(mean_logCPM = mean(logCPM, na.rm = TRUE))

ggplot(mean_expression, aes(mean_logCPM, y = Gene, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Mean Relative Expression of Genes in Each Atrium",
       x = "Gene",
       y = "Mean logCPM",
       fill = "Group") +
  facet_wrap(~ Region, scales = "free") +
  theme_classic()
```

##Curated ion-channel list
```{r}
ion_channel_list <- c(
    "SCN1A", "SCN2A", "SCN3A", "SCN4A", "SCN5A", "SCN7A", "SCN8A", "SCN9A", "SCN10A", "SCN11A",
    "KCNQ1", "KCNQ2", "KCNQ3", "KCNQ4", "KCNQ5", "KCNN1", "KCNN2", "KCNN3",
    "CACNA1C", "CACNA1D", "CACNA1F", "CACNA1S", "CACNA1A", "CACNA1B", "CACNA1E", "CACNA1H",
    "CACNA2D1", "CACNA2D2", "CACNA2D3", "CACNA2D4",
    "CACNB1", "CACNB2", "CACNB3", "CACNB4",
    "TRPC1", "TRPC2", "TRPC3", "TRPC4", "TRPC5", "TRPC6", "TRPC7",
    "TRPM1", "TRPM2", "TRPM3", "TRPM4", "TRPM5", "TRPM6", "TRPM7", "TRPM8",
    "TRPV1", "TRPV2", "TRPV3", "TRPV4", "TRPV5", "TRPV6", "CACNA1G", "KCND2", "KCND3", "KCNA4", "KCNA7", "KCNC4", 
    "KCNA5", "KCNC1", "KCNH2", "KCNQ1", "KCNJ2", "KCNJ12", "KCNJ11", 
    "KCNJ3", "KCNJ5", "KCNK1", "KCNK6", "KCNK3", "KCNK4", "HCN1", "HCN2", "HCN4",
    "SK1", "SK2", "SK3", "SLC8A1", "SLC8A2", "SLC8A3",
    "ATP2A1", "ATP2A2", "ATP2A3",
    "CASQ1", "CASQ2", "RYR1", "RYR2", "PLN",
    "GJA1", "GJA2", "GJA3", "GJA4",  "GJA5", "GJA6", "KCNA3",  "KCNA6",  "KCNB1",  "KCNC3",  "KCND1",  "KCNH8",  "KCNJ1",  "KCNJ15", "KCNJ13", "KCNJ4",  "KCNJ14", "KCNJ9",  "KCNG1",  "KCNG4")
ion_channel_list <- annot[annot$GENENAME %in% ion_channel_list, c("GENENAME", "ENSEMBL")]
ion_channel_list<- ion_channel_list[order(ion_channel_list$GENENAME), ]

#Remove missing values and create datatable
missing_ids <- setdiff(ion_channel_list$ENSEMBL, rownames(logCPM))
print(missing_ids)
ion_channel_list <- ion_channel_list[!ion_channel_list$ENSEMBL %in% missing_ids, ]
ionchannel_logCPM <- as.data.frame(logCPM[row.names(logCPM) %in% ion_channel_list$ENSEMBL, ])
ionchannel_logCPM_with_genes <- merge(ionchannel_logCPM, annot, by.x = "row.names", by.y = "ENSEMBL", all.x = TRUE)
ionchannel_melted <- data.table::melt(ionchannel_logCPM_with_genes)
setnames(x = ionchannel_melted, 
         old = c("GENENAME", "variable", "value"),
         new = c("Gene", "Sample", "logCPM"))
ionchannel_melted <- data.table::merge.data.table(x = ionchannel_melted, 
                                              y = data.table(Sample = rownames(meta), meta), 
                                              by = "Sample")

#Plot 
mean_expression <- ionchannel_melted %>% group_by(Gene, Condition, Region, Group) %>%
  summarize(mean_logCPM = mean(logCPM, na.rm = TRUE))

fwrite(x = mean_expression, 
       file = "../../../analysis/03_ion-channel/output/ion-channellist.tsv", 
       sep = "\t")
openxlsx::write.xlsx(x = mean_expression, file = "../../../analysis/03_ion-channel/output/ion-channellist.xlsx", asTable = TRUE)


ggplot(mean_expression, aes(mean_logCPM, y = Gene, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Mean Relative Expression of Genes in Each Condition",
       x = "Gene",
       y = "Mean logCPM",
       fill = "Group") +
  facet_wrap(~ Region, scales = "free") +
  theme_classic()

#Plot with * indicating significance
sig_ionchannels <- res_all[res_all$ENSEMBL %in% ion_channel_list$ENSEMBL & res_all$adj.P.Val < 0.05, ]
sig_RA_ionchannels <- sig_ionchannels[sig_ionchannels$Contrast == "AF_vs_control_RA", "GENENAME"]
sig_LA_ionchannels <- sig_ionchannels[sig_ionchannels$Contrast == "AF_vs_control_LA", "GENENAME"]
sig_RV_ionchannels <- sig_ionchannels[sig_ionchannels$Contrast == "AF_vs_control_RV", "GENENAME"]
sig_LV_ionchannels <- sig_ionchannels[sig_ionchannels$Contrast == "AF_vs_control_LV", "GENENAME"]
sig_atria_ionchannels <- sig_ionchannels[sig_ionchannels$Contrast == "Avr.Dis.Effect_Atrial", "GENENAME"]
sig_left_right_atria <- sig_ionchannels[sig_ionchannels$Contrast == "RA_vs_LA_control", "GENENAME"]
sig_left_right_ventricular <- sig_ionchannels[sig_ionchannels$Contrast == "RV_vs_LV_control", "GENENAME"]

mean_expression <- mean_expression %>%
  mutate(
    Contrast = ifelse(Region == "LA", "AF_vs_control_LA", "AF_vs_control_RA"),
    Significant = ifelse((Region == "LA" & Gene %in% sig_LA_ionchannels) |
                         (Region == "RA" & Gene %in% sig_RA_ionchannels), TRUE, FALSE)
  )

# I cant make a star to indicate significance, #TODO! 
ggplot(mean_expression, aes(mean_logCPM, y = Gene, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  geom_text(data = mean_expression %>% filter(Significant & Region == "LA"), 
            aes(label = Gene), 
            vjust = -0.5, 
            color = "black", 
            size = 4, 
            fontface = "bold") +  
  geom_text(data = mean_expression %>% filter(Significant & Region == "RA"), 
            aes(label = Gene), 
            vjust = -0.5, 
            color = "black", 
            size = 4, 
            fontface = "bold") +  
    geom_text(data = mean_expression %>% filter(Significant & Region == "LV"), 
            aes(label = Gene), 
            vjust = -0.5, 
            color = "black", 
            size = 4, 
            fontface = "bold") +  
  
  labs(title = "Mean Relative Expression of Genes in Each Condition",
       x = "Mean logCPM",
       y = "Gene",
       fill = "Group") +
  facet_wrap(~ Region, scales = "free") +
  theme_classic()

```

#Attempt to create pheatmap of ionchannel genes
```{r}
#All regions
library(pheatmap)
ion_channel_ensembl <- annot[annot$GENENAME %in% ion_channel_list$GENENAME, c("GENENAME", "ENSEMBL")]
subset_logCPM <- logCPM_batchRemoved[rownames(logCPM_batchRemoved) %in% ion_channel_ensembl$ENSEMBL, ]
rownames(subset_logCPM) <- ion_channel_ensembl$GENENAME[match(rownames(subset_logCPM), ion_channel_ensembl$ENSEMBL)]
pheatmap(subset_logCPM,
         annotation_col = meta_reordered[, c("Group", "Region")],
         scale = "row",
         show_rownames = TRUE)


#LA only sig genes
LA_subset_logCPM <- subset_logCPM[, row.names(meta)[meta$Region == "LA"]]
LA_subset_logCPM <- LA_subset_logCPM[row.names(LA_subset_logCPM) %in% sig_LA_ionchannels, ]
pheatmap(LA_subset_logCPM,
         annotation_col = meta_reordered[, c("Condition", "Group")],
         scale = "row",
         show_rownames = TRUE)

#RA only
RA_subset_logCPM <- subset_logCPM[, row.names(meta)[meta$Region == "RA"]]
RA_subset_logCPM <- RA_subset_logCPM[row.names(RA_subset_logCPM) %in% sig_RA_ionchannels, ]
pheatmap(RA_subset_logCPM,
         annotation_col = meta_reordered[, c("Condition", "Group")],
         scale = "row",
         show_rownames = TRUE)



```

#Curated gene_list_ventricular
```{r}

ventricular_channel_list <- c("CACNA1C", "CACNB2", "CACNA1D", "CACNA1G", "HCN2", "KCNJ2", "KCNJ4", "KCNJ8", "KCNJ11", "KCNH2","KCNQ1", "KCNE1", "KCNA5", "SCN5A", "ATP1A3", "SLC8A1", "KCND2", "KCND3", "KCNA4", "KCNIP2", "SERCA2A", "PLN", "RYR2")
  
ventricular_channel_list <- annot[annot$GENENAME %in% ventricular_channel_list, c("GENENAME", "ENSEMBL")]
ventricular_channel_list<- ventricular_channel_list[order(ventricular_channel_list$GENENAME), ]

#Remove missing values and create datatable
missing_ids <- setdiff(ventricular_channel_list$ENSEMBL, rownames(logCPM_batchRemoved))
print(missing_ids)
ventricular_channel_list <- ventricular_channel_list[!ventricular_channel_list$ENSEMBL %in% missing_ids, ]
ionchannel_logCPM_batchRemoved <- as.data.frame(logCPM_batchRemoved[row.names(logCPM_batchRemoved) %in% ventricular_channel_list$ENSEMBL, ])
ionchannel_logCPM_with_genes <- merge(ionchannel_logCPM, annot, by.x = "row.names", by.y = "ENSEMBL", all.x = TRUE)
ionchannel_melted <- data.table::melt(ionchannel_logCPM_with_genes)
setnames(x = ionchannel_melted, 
         old = c("GENENAME", "variable", "value"),
         new = c("Gene", "Sample", "logCPM"))
ionchannel_melted <- data.table::merge.data.table(x = ionchannel_melted, 
                                              y = data.table(Sample = rownames(meta), meta), 
                                              by = "Sample")

#Plot 
mean_expression <- ionchannel_melted %>% group_by(Gene, Condition, Region, Group) %>%
  summarize(mean_logCPM = mean(logCPM, na.rm = TRUE))

ggplot(mean_expression, aes(mean_logCPM, y = Gene, fill = Group)) +
  geom_bar(stat = "identity", position = "dodge") +
  labs(title = "Mean Relative Expression of Genes in Each Condition",
       x = "Gene",
       y = "Mean logCPM",
       fill = "Group") +
  facet_wrap(~ Region, scales = "free") +
  theme_classic()

#Heatmaps
ventricular_channel_ensembl <- annot[annot$GENENAME %in% ventricular_channel_list$GENENAME, c("GENENAME", "ENSEMBL")]
subset_logCPM <- logCPM[rownames(logCPM) %in% ventricular_channel_ensembl$ENSEMBL, ]
rownames(subset_logCPM) <- ventricular_channel_ensembl$GENENAME[match(rownames(subset_logCPM), ventricular_channel_ensembl$ENSEMBL)]
pheatmap(subset_logCPM,
         annotation_col = meta_reordered[, c("Region", "Group")],
         scale = "row",
         show_rownames = TRUE)

sig_ionchannels <- res_all[res_all$ENSEMBL %in% ventricular_channel_list$ENSEMBL & res_all$adj.P.Val < 0.05, ]
sig_RA_ionchannels <- sig_ionchannels[sig_ionchannels$Contrast == "AF_vs_control_RA", "GENENAME"]
sig_LA_ionchannels <- sig_ionchannels[sig_ionchannels$Contrast == "AF_vs_control_LA", "GENENAME"]
sig_RV_ionchannels <- sig_ionchannels[sig_ionchannels$Contrast == "AF_vs_control_RV", "GENENAME"]
sig_LV_ionchannels <- sig_ionchannels[sig_ionchannels$Contrast == "AF_vs_control_LV", "GENENAME"]
sig_atria_ionchannels <- sig_ionchannels[sig_ionchannels$Contrast == "Avr.Dis.Effect_Atrial", "GENENAME"]
sig_left_right_atria <- sig_ionchannels[sig_ionchannels$Contrast == "RA_vs_LA_control", "GENENAME"]
sig_left_right_ventricular <- sig_ionchannels[sig_ionchannels$Contrast == "RV_vs_LV_control", "GENENAME"]

sig_atrio_ventricular <- sig_ionchannels[sig_ionchannels$Contrast == "AverageRegionEffect_AV_healthy", "GENENAME"]



#RA only sig genes
RA_subset_logCPM <- subset_logCPM[, row.names(meta)[meta$Region == "RA"]]
RA_subset_logCPM <- RA_subset_logCPM[row.names(RA_subset_logCPM) %in% sig_RA_ionchannels, ]
pheatmap(RA_subset_logCPM,
         annotation_col = meta_reordered[, c("Condition", "Group")],
         scale = "row",
         show_rownames = TRUE)



```

#For Toronto Meeting
##Define function for plotting set of ion-channels in health
```{r}
library(ggplot2)
library(dplyr)
library(data.table)

#KU-red color palette
KUred <- c("#ECD9D9", "#DDBBBB", "#B05D5F", "#A6494B")
useKUred <- colorRampPalette(KUred)



plot_gene_expression <- function(gene_list, logCPM, annot, meta) {
  # Filter annotation data for the specified gene list
  gene_annotation <- annot[annot$GENENAME %in% gene_list, c("GENENAME", "ENSEMBL")]
  
  # Remove missing values
  missing_ids <- setdiff(gene_annotation$ENSEMBL, rownames(logCPM))
  gene_annotation <- gene_annotation[!gene_annotation$ENSEMBL %in% missing_ids, ]
  
  # Merge logCPM data with annotation and metadata
  gene_data <- merge(as.data.frame(logCPM_batchRemoved[row.names(logCPM_batchRemoved) %in% gene_annotation$ENSEMBL, ]), 
                     gene_annotation, by.x = "row.names", by.y = "ENSEMBL", all.x = TRUE)
  gene_melted <- data.table::melt(gene_data, id.vars = c("GENENAME"))
  setnames(gene_melted, 
           old = c("GENENAME", "variable", "value"),
           new = c("Gene", "Sample", "logCPM"))
  gene_melted <- data.table::merge.data.table(x = gene_melted, 
                                               y = data.table(Sample = rownames(meta), meta), 
                                               by = "Sample")
  gene_melted$Region <- factor(gene_melted$Region, levels = c("LA", "RA", "LV", "RV"))
  gene_melted <- gene_melted %>%
    filter(Group == "Control")
  
  gene_melted$logCPM <- as.numeric(gene_melted$logCPM)
  gene_melted <- gene_melted[!is.na(gene_melted$logCPM), ]
  
  # Calculate mean and standard deviation
  mean_expression <- gene_melted %>% 
    group_by(Gene, Region) %>% 
    summarize(mean_logCPM = mean(logCPM, na.rm = TRUE),
              sd_logCPM = sd(logCPM, na.rm = TRUE))
  
  # Plot
  ggplot() +
    geom_col(data = mean_expression, aes(x = Gene, y = mean_logCPM, fill = Region), 
             position = position_dodge(width = 0.9), alpha = 0.5) +
    geom_errorbar(data = mean_expression, aes(x = Gene, y = mean_logCPM, ymin = mean_logCPM - sd_logCPM, ymax = mean_logCPM + sd_logCPM, color = Region), 
                  position = position_dodge(width = 0.9), width = 0.25, alpha = 0.6) +
    geom_point(data = gene_melted, aes(x = Gene, y = logCPM, color = Region), 
               position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.8), alpha = 0.6) +
    labs(title = " ",
         x = " ",
         y = "Relative expression (logCPM)",
         color = "Region",
         fill = "Region") +
    theme_classic() +
    theme(legend.position = "bottom")
}

# Example usage:
gene_list <- c("CACNA1C", "CACNA1D", "CACNA1G", "CACNA1H", "CACNA2D2", "CACNA2D1")
plot_gene_expression(gene_list, logCPM, annot, meta)
```

#Example of KU red
```{r}
# Define the KUred color palette
KUred <- c("#ECD9D9", "#DDBBBB", "#B05D5F", "#A6494B")
KUalt <- c("#A6494B", "#784972", "#3C4C69", "#2F4858")
KUalt <- c("#671D24", "#B05D5F", "#574141", "#999999")


plot_gene_expression <- function(gene_list, logCPM, annot, meta) {
  # Filter annotation data for the specified gene list
  gene_annotation <- annot[annot$GENENAME %in% gene_list, c("GENENAME", "ENSEMBL")]
  
  # Remove missing values
  missing_ids <- setdiff(gene_annotation$ENSEMBL, rownames(logCPM))
  gene_annotation <- gene_annotation[!gene_annotation$ENSEMBL %in% missing_ids, ]
  
  # Merge logCPM data with annotation and metadata
  gene_data <- merge(logCPM[rownames(logCPM) %in% gene_annotation$ENSEMBL, ], 
                     gene_annotation, by.x = "row.names", by.y = "ENSEMBL", all.x = TRUE)
  gene_melted <- data.table::melt(gene_data, id.vars = c("GENENAME"))
  setnames(gene_melted, 
           old = c("GENENAME", "variable", "value"),
           new = c("Gene", "Sample", "logCPM"))
  gene_melted <- data.table::merge.data.table(x = gene_melted, 
                                               y = data.table(Sample = rownames(meta), meta), 
                                               by = "Sample")
  gene_melted$Region <- factor(gene_melted$Region, levels = c("LA", "RA", "LV", "RV"))
  gene_melted <- gene_melted %>%
    dplyr::filter(Group == "Control")
  
  gene_melted$logCPM <- as.numeric(gene_melted$logCPM)
  gene_melted <- gene_melted[!is.na(gene_melted$logCPM), ]
  
  # Calculate mean and standard deviation
  mean_expression <- gene_melted %>% 
    dplyr::group_by(Gene, Region) %>% 
    dplyr::summarize(mean_logCPM = mean(logCPM, na.rm = TRUE),
              sd_logCPM = sd(logCPM, na.rm = TRUE))
  
  # Define region-specific colors using the custom palette
  region_colors <- setNames(KUalt, levels(gene_melted$Region))
  
  # Plot
  ggplot() +
    geom_col(data = mean_expression, aes(x = Gene, y = mean_logCPM, fill = Region), 
             position = position_dodge(width = 0.9), alpha = 0.8) +
    scale_fill_manual(values = region_colors) +
    geom_errorbar(data = mean_expression, aes(x = Gene, y = mean_logCPM, ymin = mean_logCPM - sd_logCPM, ymax = mean_logCPM + sd_logCPM, color = Region), 
                  position = position_dodge(width = 0.9), width = 0.25, alpha = 0.7) +
    scale_color_manual(values = region_colors) +
    geom_point(data = gene_melted, aes(x = Gene, y = logCPM, color = Region), 
               position = position_jitterdodge(jitter.width = 0.2, dodge.width = 0.85), alpha = 0.95) +
    labs(title = "Gene Expression by Region",
         x = "Gene",
         y = "Relative expression (logCPM)",
         color = "Region",
         fill = "Region") +
    theme_classic() +
    theme(legend.position = "bottom")
}

gene_list <- c("CACNA1C", "CACNA1D", "CACNA1G", "CACNA1H", "CACNA2D2", "CACNA2D1")
plot_gene_expression(gene_list, logCPM, annot, meta)


```


##Connexins
```{r}
gene_list <- c("GJA1", "GJA2", "GJA3","GJA4", "GJA5", "GJA6", "GJC1")
plot_gene_expression(gene_list, logCPM, annot, meta)
```

##Sodium channels
```{r}
gene_list <- c("SCN1A", "SCN2A", "SCN3A", "SCN5A", "SCN9A", "SCN7A", "SCN8A", "SCN10A", "SCN1B", "SCNB3", "SCN6A", "SCN7A", "SCN4B")
plot_gene_expression(gene_list, logCPM, annot, meta)

gene_list <- c("SCN1A", "SCN2A", "SCN3A", "SCN5A", "SCN9A", "SCN7A", "SCN8A", "SCN10A", "SCN1B", "SCNB3", "SCN6A", "SCN7A", "SLC12A2", "SLC12A1")
plot_gene_expression(gene_list, logCPM, annot, meta)

```
##Voltage-gated potassium
```{r}
gene_list <- c("KCNH2", "KCNQ1", "KCND3", "KCNA4", "KCNA5")
plot_gene_expression(gene_list, logCPM, annot, meta)

```
##Indwardly rectifying potassium channels
```{r}
gene_list <- c("KCNJ3", "KCNJ5", "KCNJ12", "KCNJ2", "KCNJ4", "KCNIP2")
plot_gene_expression(gene_list, logCPM, annot, meta)

```
##SK channels
```{r}
gene_list <- c("KCNN1", "KCNN2", "KCNN3")
plot_gene_expression(gene_list, logCPM, annot, meta)

```


##HCN-channels
```{r}
gene_list <- c("HCN1", "HCN2", "HCN3", "HCN4")
plot_gene_expression(gene_list, logCPM, annot, meta)
```

#Caclium-handling and signalling
```{r}
gene_list <- c("ATP2A1", "ATP2A2", "CALM1", "CALM2", "CALM3", "CASQ1", "CASQ2", "PLB", "PLN", "RYR1", "RYR2", "INSP3R1", "INSP3R2")
plot_gene_expression(gene_list, logCPM, annot, meta)
```
#For Leguilette